---
title: "ZI-MLN: a simple simulation walkthrough"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8,
                      fig.height = 4, dpi = 100)
```

This short walkthrough fits the **ZI-MLN** model to one small, simulated data
set and checks that it recovers the truth. It focuses on the three inputs new
users ask about most: the count table `Y`, the subject index `m`, and the number
of subjects `M`.

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("shuang-jie/ZI-MLN")
```

```{r library}
library(ZIMLN)
```

## The model inputs

`ZI_MLN(Y, X = NULL, m, M, ...)` takes:

| Input | Shape | Meaning |
| --- | --- | --- |
| `Y` | `n x J` integer matrix | The **count table**. Each **row is a sample**, each **column is an OTU/feature**. Raw counts — no normalization needed, and zeros are expected and modelled. |
| `m` | length-`n` integer vector | The **subject index of each sample**: `m[i]` is the subject that sample `i` was taken from. Lets several samples share a subject (repeated measures). |
| `M` | single integer | The **number of distinct subjects**, i.e. `length(unique(m))`. Defaults to that if omitted. |
| `X` | `n x P` matrix (optional) | Sample covariates. Leave `NULL` for the no-covariate model; supply it to estimate covariate effects `beta`. |

### Understanding `m` and `M`

`m` tells the model which samples came from the same subject, so it can share a
subject-level random effect between them. A couple of common cases:

- **One sample per subject** (the simplest): `m = 1:n` and `M = n`. Every sample
  is its own subject.
- **Repeated measures**: e.g. 12 samples from 6 subjects, 2 samples each, would be
  `m = c(1, 1, 2, 2, 3, 3, 4, 4, 5, 5, 6, 6)` and `M = 6` — sample 1 and 2 share
  subject 1, samples 3 and 4 share subject 2, and so on.

The values in `m` are just labels `1, ..., M`; only which samples share a label
matters.

## Simulate a small data set

We use `simulate_zimln()` to draw a count table from the model itself, together
with the true parameters. Here we take **30 samples from `M = 10` subjects, 3
samples each**, and `J = 25` OTUs.

```{r simulate}
M <- 10
m <- rep(1:M, each = 3)            # 3 samples per subject
m
M

sim <- simulate_zimln(n = 30, J = 25, K = 3, zero.rate = 0.5,
                      M = M, m = m, seed = 1)

Y <- sim$Y
dim(Y)                             # 30 samples x 25 OTUs
Y[1:6, 1:8]                        # a corner of the count table
mean(Y == 0)                       # fraction of zeros
```

Each row of `Y` is a sample; rows sharing a subject (e.g. rows 1-3) are the
repeated measures for that subject.

## Fit the model

With the inputs assembled, one call runs the MCMC sampler. We use a fairly short
chain here so the document knits quickly; for real analyses use the defaults
(`niter = 20000`).

```{r fit}
fit <- ZI_MLN(Y, m = m, M = M, niter = 4000, burnin = 2000, seed = 7)
length(fit)                        # number of retained posterior draws
```

`fit` is a list of posterior draws. The helper functions summarise it.

## Posterior summaries and recovery

**Marginal OTU correlations** `rho_jj'` — the model's main target. We compare the
posterior mean to the truth on the off-diagonal entries:

```{r correlation}
rho_est <- posterior_correlation(fit)          # J x J matrix
ut <- upper.tri(rho_est)
cor(rho_est[ut], sim$true.cor[ut])             # agreement with the truth
```

```{r heatmap, echo = FALSE}
J <- ncol(Y)
op <- par(mfrow = c(1, 2), mar = c(2, 2, 2, 1))
image(sim$true.cor[, J:1], zlim = c(-1, 1), axes = FALSE,
      main = "True correlation", col = hcl.colors(50, "Blue-Red"))
image(rho_est[, J:1],       zlim = c(-1, 1), axes = FALSE,
      main = "Posterior mean", col = hcl.colors(50, "Blue-Red"))
par(op)
```

**Covariance** is available too — marginal `Omega` (default) or the interaction
covariance `Sigma = Lambda Lambda' + sigma^2 I`:

```{r covariance}
Sigma_est <- posterior_covariance(fit, marginal = FALSE)
dim(Sigma_est)
```

**Overall abundance** `r_i + alpha_j` (only the sum is identifiable) is recovered
from the draws:

```{r abundance}
rt_est <- Reduce(`+`, lapply(fit, function(d) outer(d$ri, d$thetaj, `+`))) / length(fit)
cor(c(rt_est), c(outer(sim$ri, sim$thetaj, `+`)))
```

## Adding covariates

To fit the covariate model, generate (or supply) an `n x P` matrix `X` and pass
it to `ZI_MLN()`. The covariate effects `beta_jp` then come from
`posterior_beta()`:

```{r covariates}
sim2 <- simulate_zimln(n = 30, J = 25, K = 3, p = 2, M = M, m = m, seed = 2)
fit2 <- ZI_MLN(sim2$Y, X = sim2$X, m = m, M = M,
               niter = 4000, burnin = 2000, seed = 7)

b <- posterior_beta(fit2)          # posterior mean + 95% intervals per (OTU, covariate)
head(b$table)
cor(c(b$mean), c(sim2$beta))       # agreement with the true beta
```

Even with these short chains the estimates track the truth; longer chains and
larger data sets give the strong recovery reported in the paper.
